Modeling and leveraging intuitive theories to improve vaccine attitudes

### 2022-02-09 paper/paper-apa-reformatting.Rmd
edits ahead of preprint

---
title             : "Modeling and leveraging intuitive theories to improve vaccine attitudes"
shorttitle        : "Intuitive theories of vaccine attitudes"

author: 
  - name          : "Derek Powell"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "4701 W. Thunderbird Road, Glendale, AZ 85306"
    email         : "dmpowell@asu.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Data collection
      - Analysis
      - Writing
  - name          : "Kara Weisman"
    affiliation   : "2"
    role:
      - Conceptualization
      - Data collection
      - Analysis
      - Writing
  - name          : "Ellen M. Markman"
    affiliation   : "3"
    role:
      - Conceptualization
      - Writing

affiliation:
  - id            : "1"
    institution   : "Arizona State University"
  - id            : "2"
    institution   : "University of California, Riverside"
  - id            : "3"
    institution   : "Stanford University"

authornote: |
  This manuscript has not yet been peer-reviewed.

abstract: |
  Much of the richness of human thought is supported by people’s intuitive theories---mental frameworks capturing the perceived structure of the world. But intuitive theories can sometimes contain and reinforce misconceptions, such as misconceptions about vaccine safety that discourage vaccination. We argue that addressing misconceptions requires awareness of the broader conceptual contexts in which they are embedded. Here, we developed a cognitive model of the intuitive theory surrounding vaccination decisions. Using this model, we were able to make accurate predictions about how people's beliefs would be revised in light of educational interventions, design an effective new intervention encouraging vaccination, and understand how these beliefs were affected by real-world events. This approach provides the foundation for richer understandings of intuitive theories and belief revision more broadly.

  <!-- https://tinyurl.com/ybremelq -->

keywords          : "keywords"
wordcount         : "X"

bibliography      : ["../supplement/references.bib"]

floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library("papaja")
r_refs("r-references.bib")
```

@@

One remarkable feature of human psychology is our ability to form rich mental representations of what is true, how things work, and how best to navigate the world around us. Much of this type of thinking can be understood as deriving from people’s *intuitive theories*---explanatory frameworks, analogous to scientific theories, that capture the causal and logical structure of the world within a domain. From early in life, people are driven to seek out such causal explanations and to use their mental representations of these causal systems to explain, predict, and intervene in the world [see @carey2009; @gelman.legare2011; @gerstenberg.etal2021; @gopnik.wellman1994; @keil1994; @murphy.medin1985; @wellman.gelman1992 for influential treatments of this topic].  Intuitive theories inform reasoning in a wide variety of domains, including reasoning about physical systems [e.g., @baillargeon2004; @battaglia.etal2013; @spelke1994a; @ullman.etal2017], biological processes [e.g., @carey1985; @carey2009;  @medin.atran1999] and the contents of other people’s minds [e.g., @baker.etal2017; @gopnik.wellman1994; @wellman2015].

@@

Even before the coronavirus pandemic, misconceptions about vaccine safety drove public health crises around the world. Undervaccination resulted in outbreaks of measles affecting over 1000 children across the United States in 2019, and over 80,000 children across Europe in 2018 [@thornton2019]. Misconceptions about childhood vaccination---particularly the fraudulent claim that vaccines cause autism or other dangerous side-effects---have proved remarkably difficult to correct [e.g., @betsch.sachse2013; @horne.etal2015]. Sadly, only a handful of educational interventions have been found to improve vaccine attitudes [e.g., @dube.etal2015; @horne.etal2015; @vannice.etal2011; @wallace.etal2006]. This is especially  sobering in the context of the coronavirus pandemic: reluctance to be vaccinated against COVID-19 threatens our ability to end a pandemic that has already claimed millions of lives around the world.

The difficulty of convincing people to vaccinate themselves and their children illustrates a general pattern of findings showing that people often do not respond to evidence in ways that educators might wish [e.g. @lewandowsky.etal2012]. Part of the challenge in correcting misconceptions is that they are situated within larger, internally coherent belief systems---intuitive theories---that guide how people interpret and respond to evidence [e.g., @gershman2019; @lewandowsky.etal2012; @jern.etal2014].

@@

Here, we sought to formalize this understanding of intuitive theories by representing an intuitive theory as a Bayesian network. Originally designed as tools for artificial intelligence [@pearl1988], Bayesian networks are a class of probabilistic graphical models that have been deployed widely across cognitive science [e.g. @gopnik.etal2004; @jern.etal2014; @griffiths.tenenbaum2005]. The heart of a Bayesian Network is a graphical structure ---not unlike a "web of belief"---that represents variables as nodes connected by edges, where those edges represent probabilistic or statistical relations.

@@

```{r fig1-maingraph, fig.align='center', out.width="90%", fig.cap="Bayesian network model resulting from hill-climbing algorithm with our generative-model constraints applied to data in Study 1 training split (n = 904). The network has 29 edges connecting these 14 variables. Edge polarity and weights are indicated by color, with red hues indicating negative or preventive influences, and blue hues indicating positive or generating influences. The direction of the edge was not determinable from the data and modeling constraints for two edges: $\\text{Naturalism} \\rightarrow \\text{Holistic Balance}$ and $\\text{Medical Skepticism} \\rightarrow \\text{Parental Expertise}$; see SI."}

@@

```{r fig2-validation, fig.align='center', out.width="100%", fig.cap="Observed responses and model predictions for each belief among participants in the Study 1 testing split. Correlations between observed and predicted values ranged from .380 to .898 across the different belief scales. In general, the model showed greater predictive accuracy for more central beliefs (e.g., \\emph{vaccine danger}) than for more peripheral beliefs (e.g., \\emph{parental protectiveness}; see SI for details)."}
knitr::include_graphics("../local/valid_plt.png")
```

@@

```{r fig3-s2, fig.align='center', out.width="100%", fig.cap="Belief changes following the intervention in Study 2. Left, mean difference scores from pretest to posttest for the 14 belief scales across experimental conditions. Right, mean observed and model predicted change scores for the disease risk intervention condition. In both plots, error bars represent 95\\% confidence intervals. In both plots, changes were calculated after transforming responses to lie between 0 and 1."}

@@

```{r fig4-predimpacts, fig.align='center',  out.width="50%", fig.cap="Predicted change in vaccination intentions following an equivalent hypothetical change in each of the other 13 beliefs."}
knitr::include_graphics("../local/pred_change_plt.png")
```

@@

```{r fig5-s3, fig.align='center',  out.width="100%", fig.cap="Belief changes following the intervention in studies 3a and 3b. Left, mean difference scores from pretest to posttest for the 14 belief scales across experimental conditions. Right, mean observed and model predicted change scores for the disease risk intervention condition. In both plots, changes were calculated after combining data from both studies and transforming responses to lie between 0 and 1 and error bars represent 95\\% confidence intervals."}
knitr::include_graphics("../local/s3_plt.png")
```

@@

```{r fig6-s4, fig.align='center',  out.width="100%", fig.cap="Participants’ belief changes from participation in Study 1 to Study 4. Left, mean difference scores comparing beliefs reported in August-September 2017 and May 2019 for the 14 belief scales, across participants who did and did not report exposure to news of the ongoing New York measles outbreak. Right, mean observed and model predicted change scores for those participants who did report exposure to this news. Belief changes were modeled assuming that news of the outbreak directly impacted participant's vaccine intentions, the most central of the 14 beliefs, in order to evaluate the overall coherence of these belief changes. In both plots, changes were calculated after transforming responses to lie between 0 and 1 and error bars represent 95\\% confidence intervals."}
knitr::include_graphics("../local/s4_plt.png")
```

@@

More broadly, the approaches we have introduced here may shed light on long-standing theoretical questions about belief revision. Many researchers have argued that people's belief updating is subject to various biases, or is skewed by motivational factors [e.g. @kahan2013; @flynn.etal2017; @kunda1990]. Yet many of the phenomena traditionally taken as evidence for these theories can also be explained by normative models of belief revision, so long as we are willing to posit that other auxiliary beliefs can affect the relationships between a belief and evidence [e.g.,@gershman2019; @jern.etal2014]. Adjudicating between competing views of belief revision will require us to move beyond minimal models and illustrative simulations by assessing whether people actually do hold these auxiliary beliefs, and whether their broader belief systems behave according to the rules of probabilistic inference. The Bayesian network structure learning techniques applied here are one tool we might use to build these richer models.

@@

On the other hand, the pandemic has presented its own unique issues. MRNA vaccines like those from Pfizer-BioNtech and Moderna represent highly novel technologiesdeveloped at a pace far faster than previously imagined possible. The unprecedented nature of these vaccines could engender hesitancy even among people who are otherwise comfortable with vaccinations.

@@

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id="refs" custom-style="Bibliography"></div>
\endgroup

>>>>

---
title: "Modeling and leveraging intuitive theories to improve vaccine attitudes"
author: "Derek Powell, Kara Weisman, and Ellen M. Markman"
output: pdf_document
bibliography: "../supplement/references.bib"
csl: "../supplement/pnas.csl"
header-includes:
    - \usepackage{setspace}\doublespacing
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

@@

# Abstract

Much of the richness of human thought is supported by people’s intuitive theories---mental frameworks capturing the perceived structure of the world. But intuitive theories can sometimes contain and reinforce misconceptions, such as misconceptions about vaccine safety that discourage vaccination. We argue that addressing misconceptions requires awareness of the broader conceptual contexts in which they are embedded. Here, we developed a cognitive model of the intuitive theory surrounding vaccination decisions. Using this model, we were able to make accurate predictions about how people's beliefs would be revised in light of educational interventions, design an effective new intervention encouraging vaccination, and understand how these beliefs were affected by real-world events. This approach provides the foundation for richer understandings of intuitive theories and belief revision more broadly.

\newpage

# Main text

One remarkable feature of human psychology is our ability to form rich mental representations of what is true, how things work, and how best to navigate the world around us. Much of this type of thinking can be understood as deriving from people’s *intuitive theories*---explanatory frameworks, analogous to scientific theories, that capture the causal and logical structure of the world within a domain. From early in life, people are driven to seek out such causal explanations and to use their mental representations of these causal systems to explain, predict, and intervene in the world [see @carey2009; @gelman.legare2011; @gerstenberg.etal2021; @gopnik.wellman1994; @keil1994; @murphy.medin1985; @wellman.gelman1992 for influential treatments of this topic].  Intuitive theories inform reasoning in a wide variety of domains, including reasoning about physical systems [e.g., @baillargeon2004; @battaglia.etal2013; @spelke1994a; @ullman.etal2017], biological processes [e.g., @carey2009; @carey1985; @medin.atran1999] and the contents of other people’s minds [e.g., @gopnik.wellman1994; @wellman2015; @baker.etal2017].

@@

Even before the coronavirus pandemic, misconceptions about vaccine safety drove public health crises around the world. Undervaccination resulted in outbreaks of measles affecting over 1000 children across the United States in 2019, and over 80,000 children across Europe in 2018 [@thornton2019]. Misconceptions about childhood vaccination---particularly the fraudulent claim that vaccines cause autism or other dangerous side-effects---have proved remarkably difficult to correct [e.g., @betsch.sachse2013; @horne.etal2015]. Sadly, only a handful of educational interventions have been found to improve vaccine attitudes [e.g., @horne.etal2015; @vannice.etal2011; @wallace.etal2006; @dube.etal2015]. This is especially  sobering in the context of the coronavirus pandemic: reluctance to be vaccinated against COVID-19 threatens our ability to end a pandemic that has already claimed millions of lives around the world.

The difficulty of convincing people to vaccinate themselves and their children illustrates a general pattern of findings showing that people often do not respond to evidence in ways that educators might wish [e.g. @lewandowsky.etal2012]. Part of the challenge in correcting misconceptions is that they are situated within larger, internally coherent belief systems---intuitive theories---that guide how people interpret and respond to evidence [e.g., @lewandowsky.etal2012; @jern.etal2014; @gershman2019].

@@

Here, we sought to formalize this understanding of intuitive theories by representing an intuitive theory as a Bayesian network. Originally designed as tools for artificial intelligence [@pearl1988], Bayesian networks are a class of probabilistic graphical models that have been deployed widely across cognitive science [e.g. @jern.etal2014; @gopnik.etal2004; @griffiths.tenenbaum2005]. The heart of a Bayesian Network is a graphical structure ---not unlike a "web of belief"---that represents variables as nodes connected by edges, where those edges represent probabilistic or statistical relations.

@@

```{r fig1-maingraph, fig.align='center', fig.height=8, fig.width=5, fig.cap="Bayesian network model resulting from hill-climbing algorithm with our generative-model constraints applied to data in Study 1 training split (n = 904). The network has 29 edges connecting these 14 variables. Edge polarity and weights are indicated by color, with red hues indicating negative or preventive influences, and blue hues indicating positive or generating influences. The direction of the edge was not determinable from the data and modeling constraints for two edges: $\\text{Naturalism} \\rightarrow \\text{Holistic Balance}$ and $\\text{Medical Skepticism} \\rightarrow \\text{Parental Expertise}$; see SI."}

@@

```{r fig2-validation, fig.align='center', fig.height=4, fig.width=7, fig.cap="Observed responses and model predictions for each belief among participants in the Study 1 testing split. Correlations between observed and predicted values ranged from .380 to .898 across the different belief scales. In general, the model showed greater predictive accuracy for more central beliefs (e.g., \\emph{vaccine danger}) than for more peripheral beliefs (e.g., \\emph{parental protectiveness}; see SI for details)."}
knitr::include_graphics("../local/valid_plt.png")
```

@@

```{r fig3-s2, fig.align='center', fig.height=4, fig.width=7, fig.cap="Belief changes following the intervention in Study 2. Left, mean difference scores from pretest to posttest for the 14 belief scales across experimental conditions. Right, mean observed and model predicted change scores for the disease risk intervention condition. In both plots, error bars represent 95\\% confidence intervals. In both plots, changes were calculated after transforming responses to lie between 0 and 1."}

@@

```{r fig4-predimpacts, fig.align='center', fig.height=3, fig.width=3, fig.cap="Predicted change in vaccination intentions following an equivalent hypothetical change in each of the other 13 beliefs."}
knitr::include_graphics("../local/pred_change_plt.png")
```

@@

```{r fig5-s3, fig.align='center', fig.height=4, fig.width=7, fig.cap="Belief changes following the intervention in studies 3a and 3b. Left, mean difference scores from pretest to posttest for the 14 belief scales across experimental conditions. Right, mean observed and model predicted change scores for the disease risk intervention condition. In both plots, changes were calculated after combining data from both studies and transforming responses to lie between 0 and 1 and error bars represent 95\\% confidence intervals."}
knitr::include_graphics("../local/s3_plt.png")
```

@@

```{r fig6-s4, fig.align='center', fig.height=4, fig.width=7, fig.cap="Participants’ belief changes from participation in Study 1 to Study 4. Left, mean difference scores comparing beliefs reported in August-September 2017 and May 2019 for the 14 belief scales, across participants who did and did not report exposure to news of the ongoing New York measles outbreak. Right, mean observed and model predicted change scores for those participants who did report exposure to this news. Belief changes were modeled assuming that news of the outbreak directly impacted participant's vaccine intentions, the most central of the 14 beliefs, in order to evaluate the overall coherence of these belief changes. In both plots, changes were calculated after transforming responses to lie between 0 and 1 and error bars represent 95\\% confidence intervals."}
knitr::include_graphics("../local/s4_plt.png")
```

@@

More broadly, the approaches we have introduced here may shed light on long-standing theoretical questions about belief revision. Many researchers have argued that people's belief updating is subject to various biases, or is skewed by motivational factors [e.g. @kahan2013; @flynn.etal2017; @kunda1990]. Yet many of the phenomena traditionally taken as evidence for these theories can also be explained by normative models of belief revision, so long as we are willing to posit that other auxiliary beliefs can affect the relationships between a belief and evidence [e.g., @jern.etal2014; @gershman2019]. Adjudicating between competing views of belief revision will require us to move beyond minimal models and illustrative simulations by assessing whether people actually do hold these auxiliary beliefs, and whether their broader belief systems behave according to the rules of probabilistic inference. The Bayesian network structure learning techniques applied here are one tool we might use to build these richer models.

@@

On the other hand, the pandemic has presented its own unique issues. MRNA vaccines like those from Pfizer-BioNtech and Moderna represent highly novel technologies that are currently approved only under emergency use authorizations after being developed at a pace far faster than previously imagined possible. These facts could engender hesitancy even among people who are otherwise comfortable with vaccinations.

@@
